---
title: "Advent of Code Day 1"
author: Henry Ngo
date: December 2, 2023
output: html_notebook
---

# Part 1 Problem description

> The newly-improved calibration document consists of lines of text; each line originally contained a specific calibration value that the Elves now need to recover. On each line, the calibration value can be found by combining the first digit and the last digit (in that order) to form a single two-digit number.
>
> For example:

```{verbatim}
1abc
pqr3stu8vwx
a1b2c3d4e5f
treb7uchet
```

> In this example, the calibration values of these four lines are 12, 38, 15, and 77. Adding these together produces 142.
>
> Consider your entire calibration document. What is the sum of all of the calibration values?

# Part 1 Solution attempt

## Load libraries

```{r load libraries}
library(here)
library(tidyverse)
library(stringi)
```

## Load input data

Need `col_names` to avoid reading first line as column headers

```{r}
cal_df <- readr::read_csv(here::here("day01","input.txt"), col_names="input")
```

Examine data

```{r}
head(cal_df)
```

## Get first and last digit

Store in new columns

```{r}
cal_df <- cal_df |>
  dplyr::mutate(
    first = stringr::str_extract(input, "[0-9]"),
    last  = stringr::str_extract(stringi::stri_reverse(input), "[0-9]")
  )
```

### Check output.

Visual inspection

```{r}
cal_df
```

Check for missing

```{r}
cal_df |> dplyr::summarise(input_sum = sum(is.na(input)),
                           first_sum = sum(is.na(first)),
                           last_sum  = sum(is.na(last)))
```

All zeroes, looks good.

## Create the two-digit number

```{r}
cal_df <- cal_df |> 
  tidyr::unite(cal_1, first, last, sep="") |>
  dplyr::mutate(cal_1 = as.numeric(cal_1))
```

## Print the sum

```{r}
cal_df$cal_1 |> sum()
```

# Part 2 Problem description

> Your calculation isn't quite right. It looks like some of the digits are actually spelled out with letters: one, two, three, four, five, six, seven, eight, and nine also count as valid "digits".
>
> Equipped with this new information, you now need to find the real first and last digit on each line. For example:

```{verbatim}
two1nine
eightwothree
abcone2threexyz
xtwone3four
4nineeightseven2
zoneight234
7pqrstsixteen
```

> In this example, the calibration values are 29, 83, 13, 24, 42, 14, and 76. Adding these together produces 281.
>
> What is the sum of all of the calibration values?

# Part 2 Solution attempt

## Replace written numbers to digit character
Need to account for cases like `8seveneightwox`, should resolve to `878wox` not `8782x`. So we can't use `str_replace_all`, which finds all the matches and then replaces them all at once. We could use `str_replace` to do the first match and then repeat But how do we know how many times to repeat without brute force?

### Attempt 1 (failed)
This was the attempt with `str_replace_all` which ran but did not handle text strings where two numbers shared a letter correctly (e.g. the `8seveneightwox` example above). So next is to try using `str_replace()`
```{r eval=FALSE}
# cal_df <- cal_df |> 
#   mutate(input2 = stringr::str_replace_all(input,c(
#     "one" = "1",
#     "two" = "2",
#     "three" = "3",
#     "four" = "4",
#     "five" = "5",
#     "six" = "6",
#     "seven" = "7",
#     "eight" = "8",
#     "nine" = "9"
#   )))
```

### Define matching and replacement patterns
Write a pattern for `stringr::str_replace()` to use for matching and a function using `base::switch()` to tell `str_replace` how to replace the matches it finds.
```{r}
match_pat <- "(one|two|three|four|five|six|seven|eight|nine)"

replacement_fn <- function(matched_text_to_replace){
  switch(matched_text_to_replace,
         one = "1",
         two = "2",
         three = "3",
         four = "4",
         five = "5",
         six = "6",
         seven = "7",
         eight = "8",
         nine = "9")
}
```


### Attempt 2 (failed)
Try to define a function that would check for the largest number of matches and then repeat the process N times. 

I think the general idea works (works on a single string) but didn't work when applying to the full data frame

```{r eval=FALSE}
# replace_num_text_seq <- function(my_str_vec){
# 
#   # Define a function to act on an individual row
#   replace_per_row <- function(input){
#     # Find max number of matches in this row
#     max_steps <- max(stringr::str_count(input, match_pat))
#     # Apply replacement function that many times
#     for(ii in 1:max_steps){
#       input <- stringr::str_replace(input, match_pat, replacement_fn)
#     }
#     return(input)
#   }
#   
#   # Apply to all the rows
#   out <- lapply(my_str_vec, replace_per_row)
#   
#   return(out)
# }
# 
# cal_df <- cal_df |> 
#   mutate(input2 = replace_num_text_seq(input))
```
### Attempt 3 (failed)
Using what we learned above, implement serially as nested for loops. Not pretty, but time to move on!

```{r}
# input_tmp <- cal_df$input
# # Loop through each vector entry individually
# for(ii in 1:length(input_tmp)){
#   #print(ii)
#   input <- input_tmp[ii]
#   # Find the maximum number of matches in this entry
#   max_steps <- max(stringr::str_count(input, match_pat))
#   # Go to next iteration if no matches
#   if(max_steps==0) next
#   # Loop through each replacement in this entry
#   for(jj in 1:max_steps){
#     input <- stringr::str_replace(input, match_pat, replacement_fn)
#   }
#   # Update vector entry
#   input_tmp[ii] <- input
# }
# cal_df$input2 <- input_tmp
```

### Attempt 4
Thanks to people on BlueSky, learned that I'm overthinking the edge cases like `oneight`. String replacement is not needed! 

We just need the first and last digit, whether it's a digit or the number spelled out. So we just need to modify the Part 1 solution to include the text form of numbers

#### Define match pattern
```{r}
match_pat2 <-  "([1-9]|one|two|three|four|five|six|seven|eight|nine)"
match_pat2_rev <- "([1-9]|eno|owt|eerht|ruof|evif|xis|neves|thgie|enin)"
```

#### Get first and last
```{r}
cal_df <- cal_df |>
  dplyr::mutate(
    first2 = stringr::str_extract(input, match_pat2),
    last2  = stringr::str_extract(stringi::stri_reverse(input), match_pat2_rev)
  )
```

### Check output.

Visual inspection

```{r}
cal_df
```

Check for missing

```{r}
cal_df |> dplyr::summarise(first_sum2 = sum(is.na(first2)),
                           last_sum2  = sum(is.na(last2)))
```

All zeroes, looks good.

## Create the two-digit number
First, we have to handle the text. Convert the spelled-out numbers as well as the character form into a numeric digit.

### Convert char digit and text to numeric
```{r}
cal_df <- cal_df |>
  dplyr::mutate(
    first2 = dplyr::case_when(
      first2 %in% c("1", "one") ~ 1,
      first2 %in% c("2", "two") ~ 2,
      first2 %in% c("3", "three") ~ 3,
      first2 %in% c("4", "four") ~ 4,
      first2 %in% c("5", "five") ~ 5,
      first2 %in% c("6", "six") ~ 6,
      first2 %in% c("7", "seven") ~ 7,
      first2 %in% c("8", "eight") ~ 8,
      first2 %in% c("9", "nine") ~ 9
    ),
    last2 = dplyr::case_when(
      last2 %in% c("1", "eno") ~ 1,
      last2 %in% c("2", "owt") ~ 2,
      last2 %in% c("3", "eerht") ~ 3,
      last2 %in% c("4", "ruof") ~ 4,
      last2 %in% c("5", "evif") ~ 5,
      last2 %in% c("6", "xis") ~ 6,
      last2 %in% c("7", "neves") ~ 7,
      last2 %in% c("8", "thgie") ~ 8,
      last2 %in% c("9", "enin") ~ 9
    )
  )
```


```{r}
cal_df <- cal_df |> 
  tidyr::unite(cal_2, first2, last2, sep="") |>
  dplyr::mutate(cal_2 = as.numeric(cal_2))
```

## Print the sum

```{r}
cal_df$cal_2 |> sum()
```